### 2014-07-17 Thesis - 2/Paper/paper.tex
Schrijven van de paper

%% if you use PostScript figures in your article
%% use the graphics package for simple commands
%% \usepackage{graphics}
%% or use the graphicx package for more complicated commands
%% \usepackage{graphicx}
%% or use the epsfig package if you prefer to use the old commands
%% \usepackage{epsfig}

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

@@

\title{Consistency comparison of HBase and MongoDB}

@@

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword

%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)

@@

\section{title}

%% The Appendices part is started with the command \appendix;
%% appendix sections are then done as normal sections
%% \appendix

@@

\bibliographystyle{elsarticle-num}
\bibliography{<your-bib-database>}

%% Authors are advised to submit their bibtex database files. They are
%% requested to list a bibtex style file in the manuscript if they do
%% not want to use elsarticle-num.bst.

>>>>

%% if you use PostScript figures in your article
%% use the graphics package for simple commands
%% \usepackage{graphics}
%% or use the graphicx package for more complicated commands
%% \usepackage{graphicx}
%% or use the epsfig package if you prefer to use the old commands
%% \usepackage{epsfig}
\usepackage{todonotes}
%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
%% \usepackage{amsthm}

@@

\title{CAP in practice: HBase}

@@

\begin{keyword}
%% keywords here, in the form: keyword \sep keyword
%% MSC codes here, in the form: \MSC code \sep code
%% or \MSC[2008] code \sep code (2000 is the default)

@@

In the end many choices can be made, but how does it reflect to the behaviour in practice? This paper will present result from experiments with HBase and MongoDB compared to the CAP theorem. The paper is organised as follows. Sect\todo{Do this}

\section{The CAP Theorem\cite{brewer2000towards}\cite{brewer2012cap}}
The CAP Theorem was introduced by E. Brewer \cite{brewer2000towards}  in 2000 and discusses 3 properties of which each network shared-data system can guarantee at most 2: (definitions based on \cite{brewer2012cap})
\begin{itemize}
\item (Strict)\textbf{C}onsistency: The system acts like there is only single storage 
\item High \textbf{a}vailability: The system is available (for updates)
\item \textbf{P}artition tolerance: A split in the network let the different partitions still act as a single system. 
\end{itemize}
Designers used this model to explain their design decisions, others used it to compare different system but sometimes it was misused. As E. Brewer explains 12 year after the launch, the "2 of 3" can be misleading.

One of the reasons is that there exist several types of consistency, partition tolerance and different availability guarantees. These choices need to be made several times in the different subsystems and the end solution is not black or white.

At first glance, it also looks like partition tolerance has to be implemented and therefore consistency and availability needs to be given up. In the practice, partition splits are only rare and therefore both consistency and availability can be allowed most of the time.

Each of the 3 choices will be discussed in more detail, how their implementation could work, what the influences are and some examples.

\subsection{CA}
When forfeiting partition tolerance, these systems provide all the time consistent data available to all nodes, except when there are one or more nodes unavailable. In that case, write requests will be not allowed.

These systems can be build around the 2 phase commit and have cache invalidation protocols. Examples of this types are the typical relational databases roll out in clusters.

\subsection{CP}
A consistent system with partition tolerance will provide all the time the last data, even in the case of network splits. This comes with the loss of all nodes available all the time.

The system will allow operations only on the majority partition. In case multiple splits are present and no partition has a majority of nodes, the whole system can be unavailable. These systems can be build around a master/slave principle where the operations will be directed to the master, the slaves are present to continue operation when the master fails.

In practice, systems like MongoDB, HBase and Redis chooses for CP.

\subsection{AP}
In a highly available systems with partition tolerance, is it possible to read inconsistent data. As read and write operations are still allowed when there are different partitions, it is possible that the database has other content depending on the used node. When the split is dissolved, a need for manual conflict resolution can be needed. In case a record is adapted in both partitions, the user will need to choose the correct version.

Example systems following AP are Cassandra, Riak and Voldemort.

\section{Overview of HBase and MongoDB}
In this article, 2 CP systems will be discussed more in detail regarding their choices to forfeit availability and the influence on their behaviour in practice. The systems are HBase and MongoDB, an architectural overview will be given in this section.

\subsection{HBase}
HBase\cite{hbase-doc} is an open-sourced, distributed, versioned database designed after Google's BigTable \cite{chang2008bigtable}. HBase relies on Zookeeper for the distributed task coordination and the persistent storage can be done on the local hard disc, Hadoop Distributed File System or Amazon S3. In this installation is chosen for Hadoop.

HBase nodes exists out of HMasters and HRegionServers, the coordination of the system is done by one HMaster, the handling of data is done by the HRegionServer. To store the data, on each HRegionServer a Hadoop datanode instance should be deployed for data storage, the data will be replicated to a configurable amount of other nodes. The data is stored in a table, which is split in one or more regions. A region is leased to a given HRegionServer for a defined time. During this time, only this server will provide the data of the region to the different users. This way the consistency of data can be guaranteed because there is for each record only a single system responsible. Consistency on a single record is provided by a readers/writer lock on a single record for the according queries, this way there is a guarantee to atomicity on a single record, the full procedure is explained by Lars Hofhansl\cite{hbase-acid}.

To be partition tolerant, the partition with the majority of the Zookeeper servers and a HMaster will appoint regions to available HRegionServers for them, let's call this the data serving partition. With this approach, it is important to place the Zookeeper and HMaster servers in diverse location as otherwise a partition of only this servers will make the whole system unavailable.

In HBase, a node will be able to answer to requests if the node is present in the data serving partition. Only in rare cases that all data copies of Hadoop are stored in the other, inactive cluster, the data will be unreachable. The nodes not in the data serving partition, will be unable to complete any requests.

When a server goes down, he can release the lease in case there is a graceful shut down (the HBase server is notified) and another server can get the lease immediately. In other cases, a new lease can only be given after the decay of the old lease, if the server comes back on-line in meantime, he will still be responsible.

\subsection{MongoDB}
MongoDB\cite{mongodb-doc} is an open-sourced, distributed database designed for document storage, this are data entries where the format of each record can be different. According to their website they provide high performance and high availability, but this is incorrect to the given definition in this article.

MongoDB provides data replication and data distribution, the first is done by grouping different MongoDB servers into a ReplicaSet, the second is done by grouping different of these ReplicaSets.

A ReplicaSet exists out of different MongoDB server which work as a master/slave configuration. A master is a primary and the slave is called a secondary. The primary is responsible for the write transactions, by default a query will succeed once it has a confirmation that the write has been executed on the primary. The read operation will go by default on the primary as well. Both query methods are configurable to give other guarantees, for a write operation there are multiple \textit{write concerns}, it is possible to wait till it has written on hard disc or a number of secondary servers, however all need a primary. For read operation there are multiple \textit{read preferences}, it is possible to read from a secondary or the closest server. \\ 
In the default configuration, there will be consistency guarantee.

In a ReplicaSet, for the primary election there is at least half of the ReplicaSet needed. As there is always a primary needed, the system has partition tolerance but no high availability, contrary to the statement on the documentation. However, it is possible to read from a secondary but writing is not possible.

The state of the different members of a ReplicaSet is maintained by a heartbeat system: a server is marked as off-line if no beat has been received for 10 seconds. In case the primary goes off-line, election will be started to re-elect a new one. In other words, a primary has a lease of 10 seconds. This value of 10 seconds is non-configurable.

De data distributions happens by merging replica sets in a cluster. Furthermore, there is the need for access server (as many as you want) and configuration servers (1 or 3). The availability and partition tolerance of the data is the same as in the ReplicaSets as it handled by the ReplicaSets. In case a access server can't reach a primary, another access server will be needed to write data, in case a majority of the configuration servers are not reachable, their will be no reconfiguration of the data over the different servers.

\subsection{Differences between databases} 
Both systems provide consistency and partition tolerance and forfeit high availability, but some differences are in their implementation.

First of all in their partition tolerance, in HBase there are dedicated management servers (HMaster and Zookeeper) to distribute the responsibilities of regions and if the management servers are in a partition with a minority of the data servers, data will be not available. In MongoDB the management of a ReplicaSet is done internally and the write queries will be available in the partition with the majority of the data servers.

Both decision has their influence, in HBase it is possible to write every record, as a new region can be created if the old region is unavailable. In MongoDB it is possible that in sharding you can read all the data from multiple secondaries but only write given ranges of records.

In availability, there is a small difference: where in MongoDB it is possible to read from secondaries, this is impossible in HBase.

In section \ref{sec:result}, a detailed analyse will be done to the consistency and the behaviour of the systems in the case of network or server failure.

\section{Test method}
To test the behaviour of database systems towards consistency and availability, the Yahou Cloud Serving Benchmarking (YCSB \cite{cooper2010benchmarking}) has been extended with event support for availability and reader-writers for consistency.

Each of this tests follows the same steps: calibrating the system, records are preloaded, the test is started and in the end all the records are removed again and they are executed for HBase and MongoDB. During the calibration, a workload is chosen so there is a medium load on the different databases.

\subsection{Event support}
The implementation of event support integrates a way to execute given UNIX commandos at specified moments, the execution time and result code is logged.

In these tests were 3 kinds of tests executed of which each one took 900 second. The first action takes place at 300 seconds, the second at 600 seconds.
\begin{itemize}
\item Graceful shut down of a data service and restart of the service
\item Hard kill of the data service and a restart of the service
\item Blocking of all network traffic from and to a server and the allowance of all network traffic
\end{itemize}

Each one tests the behaviour of the system under different circumstances, the first two checks what happens in case of a respectively planned and crash shut down, the latter check the behaviour in case the network fails.

These tests are executed on each of the data nodes of HBase and MongoDB.

\section{Results}\label{sec:result}
(50 threads, 600 queries/second) and MongoDB (15 threads, 200 queries/second). These values were based 
\section{Related work}

\section{Conclusion}
%% The Appendices part is started with the command \appendix;
%% appendix sections are then done as normal sections
%% \appendix

@@

\bibliographystyle{elsarticle-num}
\bibliography{referenties}
%% Authors are advised to submit their bibtex database files. They are
%% requested to list a bibtex style file in the manuscript if they do
%% not want to use elsarticle-num.bst.

### 2014-07-17 Thesis - 2/appendix/app-IMP.tex
Schrijven van de paper

\chapter{Uitwerking IMP} \label{chap:AppendixUitwerkingIMP}

>>>>

\chapter{Uitwerking IMP} \label{chap:AppendixUitwerkingIMP}
In dit hoofdstuk zal voor de verschillende systemen de automatisering van installatie en configuratie met behulp van IMP uitgelegd worden. Er zal steeds de afhankelijkheden gegeven worden, een domeinmodel, uitleg bij het domeinmodel en voorbeeld configuratie gegeven worden.

De automatisatie van installatie is ontwikkeld en getest met Fedora 18 en 20, op andere distributies en versies is er niet getest. 
Elk systeem maakt gebruik van \textit{ip::services::Server}, een instantie hiervan is een (virtuele) machine met een IP adres en besturingssysteem.

Bij elke instantie is het verplicht om de firewall uit te zetten en SELinux op permissive te zetten. Dit kan met behulp van de volgende commando's: 
\begin{lstlisting}[frame=single, breaklines=true]
systemctl stop firewalld.service  
systemctl disable firewalld.service  
setenforce 0
sed -i "s/SELINUX=enforcing/SELINUX=permissive/g" /etc/sysconfig/selinux
sed -i "s/SELINUX=enforcing/SELINUX=permissive/g" /etc/selinux/config|
\end{lstlisting}

\section{HBase}
\textit{Link: \url{https://github.com/thuys/hbase}}

Benodigde IMP modules: std, net, ip, redhat, hosts en yum.

De installatie en configuratie is gebeurd aan de hand van de uitleg en yum-repository van Cloudera\footnote{\url{http://www.cloudera.com/content/cloudera-content/cloudera-docs/CDH4/4.2.0/CDH4-Installation-Guide/CDH4-Installation-Guide.html}}.

\subsection{Domein model en uitleg}
Het domeinmodel is te zien in figuur \ref{fig:imp-hbase-domeinmodel}.

\paragraph{HBaseMaster} Dit is de implementatie van de HMaster, dient toegewezen worden aan een host met java installatie. De poort is de poort waarop de HMaster actief is.

\paragraph{HRegion} Dit is de implementatie van de HRegionServer, dient toegewezen worden aan een host met java installatie. De poort is de poort waarop de HRegionServer actief is.

\paragraph{HadoopHDFS} Dit is de implementatie van de HDFS namenode, dient toegewezen worden aan een host met java installatie. De poort is de poort waarop de namenode actief is, de directory is de directory voor HBase  en de nameDir de lokatie waar de data op harde schijf weggeschreven zal worden.

\paragraph{HadooDatapHDFS} Dit is de implementatie van de HDFS datanode, dient toegewezen worden aan een host met java installatie. De poort is de poort waarop de namenode datanode is, de directory is de directory voor HBase en de dataDir de lokatie waar de data op harde schijf weggeschreven zal worden.

\paragraph{Zookeeper} Dit is de implementatie van een enkele Zookeeper. Bij het toewijzen van meerdere aan een cluster zullen de Zookeepers een cluster vormen.

\paragraph{Javahost} Dit is een server waar Java is op geïnstalleerd.

\begin{figure}[ht!]
\centering
\includegraphics[width=\linewidth]{img/HBase-Domeinmodel.png}
\caption{HBase: Domeinmodel HBase in IMP}
\label{fig:imp-hbase-domeinmodel}
\end{figure}

\subsection{Voorbeeld configuratie}
De configuratie voor de testomgeving gaat als volgt:

\lstinputlisting[language=Python, breaklines=true, frame=single]{code/imp-hbase.conf}

\section{MongoDB}
\textit{Link: \url{https://github.com/thuys/mongodb}}

Benodigde IMP modules: std, net, ip, redhat, hosts en yum.

De installatie en configuratie is gebeurd aan de hand van de uitleg en yum-repository van MongoDB\footnote{\url {http://docs.mongodb.org/manual/tutorial/install-mongodb- on-red-hat-centos-or-fedora-linux/}, \url{http://docs.mongodb.org/manual/tutorial/deploy-replica-set-for-testing/} en  \url{http://docs.mongodb.org/manual/tutorial/deploy-shard-cluster/}}.

\subsection{Domein model en uitleg}
Het domeinmodel is te zien in figuur \ref{fig:imp-mongodb-domeinmodel}.

	\paragraph{MongoDB} is een server in het IMP model en is verantwoordelijk voor het installeren van de basis van MongoDB. Hierna zijn basis commando's voor connectie te maken met een MongoDB instantie beschikbaar. 
	\paragraph{MonogDBServer} is een server in het IMP model en is verantwoordelijk voor het installeren van de MongoDB server. 
	\paragraph{MongoDBNode} is de implementatie van een data instantie, maximaal 1 per server. Indien gelinkt met een replica set zal deze als een deel van een replica set worden geïnitialiseerd, anders als een zelfstandige instantie. 
	\paragraph{MongoDBReplicaSet} is de voorstelling van een replica set, dit wordt niet aan een specifieke server toegewezen. 
	\paragraph{MongoDBReplicaSetController} is verantwoordelijk om de replica set te initialiseren. Belangrijk is dat indien er een uitbreiding is van de set, de node verbonden met de controller een reeds geïnitialiseerde node is.  
	\paragraph{MongoDBConfigServer} is de implementatie van een configuratie server, 1 of 3 servers zijn nodig per cluster. 
	\paragraph{MongoDBAccessServer} is de implementatie van mongos, minstens 1 is nodig maar meer kunnen gebruikt worden.
	\paragraph{MongoDBShardCluster} is de voorstelling van een cluster van shards, er kunnen zowel alleenstaand instanties als replica sets aan toegevoegd worden. 
	\paragraph{MongoDBShardController} is verantwoordelijk om de cluster te initialiseren met de verschillende shards, databases, collecties en keys. 
	\paragraph{MongoDBDatabase} is de voorstelling van een database.
	\paragraph{MongoDBCollection} is de voorstelling van een collectie, indien verbonden met een cluster via een database zal deze gedeeld worden over de verschillende shards. 
	\paragraph{MongoDBKey} is de wijze waarmee een collectie verdeeld wordt over de verschillende shards.

\begin{figure}[ht!]
\centering
\includegraphics[width=\linewidth]{img/MongoDB-Domeinmodel.png}
\caption{MongoDB: Domeinmodel MongoDB in IMP}
\label{fig:imp-mongodb-domeinmodel}
\end{figure}

\subsection{Voorbeeld configuratie}
De configuratie voor de testomgeving gaat als onderstaand. Bij de uitrol van IMP gaat dit verschillende keren uitgevoerd moeten worden omdat eerst de MongoDBNodes moeten draaien, vervolgens kunnen de replicasets aangemaakt worden, daarna kunnen de replicasets pas toegevoegd worden in de cluster.

In IMP was het nog niet mogelijk om een te zeggen dat x uitgevoerd moet zijn op een andere instantie, vooraleer y kan uitgevoerd worden, ondertussen is dit mogelijk door de thesis van Harm De Weirdt\cite{thesisHarm} waar de nieuwe installatie beschikbaar is op \url{https://github.com/Foezjie/mongodb} maar hierbij dient ook gebruik gemaakt te worden van zijn IMP installatie.  \\
Met het ontbreken hieraan kan het zijn dat er 3 keer een volledige IMP deploy uitgevoerd moet worden.

\lstinputlisting[language=Python, breaklines=true, frame=single]{code/imp-mongodb.conf}

\section{Pgpool-II}
\textit{Link: \url{https://github.com/thuys/postgresql}}

Benodigde IMP modules: std, net, ip, redhat, hosts en yum.

De installatie en configuratie is gebeurd aan de hand van de uitleg Pgpool-II\footnote{\url {http://pgpool.projects.pgfoundry.org/pgpool-II/doc/tutorial-en.html/}}.

\subsection{Domein model en uitleg}
Het domeinmodel is te zien in figuur \ref{fig:imp-pgpool-domeinmodel}.

	\paragraph{PgpoolMain} Dit is de implementatie van de Pgpool-II router node.

	\paragraph{PgpoolNode} Dit is de implementatie van de Pgpool-II data node die een uitbreiding is van de standaard PostgreSQL installatie.

	\paragraph{PostgresqlServer} Dit is de implementatie van de standalone PostgreSQL server.

\begin{figure}[ht!]
\centering
\includegraphics[width=\linewidth]{img/Postgres-Domeinmodel.png}
\caption{Pgpool-II: Domeinmodel Pgpool-II in IMP}
\label{fig:imp-pgpool-domeinmodel}
\end{figure}

\subsection{Voorbeeld configuratie}
De configuratie van Pgpool-II gebeurt in verschillende stappen: shell code, IMP uitrol, extra configuratie stap, IMP uitrol.

De eerste shell code bestaat erin om de SELinux volledig uit te schakelen: 
\begin{lstlisting}[frame=single, breaklines=true]
systemctl stop firewalld.service  
systemctl disable firewalld.service  
echo "SELINUX=disabled SELINUXTYPE=targeted" > /etc/selinux/config

echo "SELINUX=disabled SELINUXTYPE=targeted" > /etc/sysconfig/selinux
\end{lstlisting}

De configuratie voor de testomgeving gaat als volgt in IMP:

\lstinputlisting[language=Python, breaklines=true, frame=single]{code/imp-pgpool.conf}

De configuratie bestaat erin om al de verschillende nodes van Pgpool-II, ongeachte of dit routers of datanodes zijn, ssh toegang te geven tot elkaar server via ssh met root en postgres als gebruikers. Deze verbinding al een keer gemaakt zijn want een bericht dat de sleutel nu mee is opgeslagen is voldoende om de online recovery te doen falen.

Hierna kan de IMP uitrol nog een keer gebeuren en het systeem zou moeten werken.

\section{YCSB}
\textit{Link: \url{https://github.com/thuys/ycsb}}

Benodigde IMP modules: std, net, ip, redhat, hosts, yum, git, hbase, mongodb, pgpool-II

De installatie en configuratie is gebeurd aan de hand van de uitleg van YCSB\footnote{\url {https://github.com/brianfrankcooper/YCSB/wiki/}}.

\subsection{Domein model en uitleg}
Het domeinmodel is te zien in figuur \ref{fig:imp-ycsb-domeinmodel}.
\begin{figure}[ht!]
\centering
\includegraphics[width=\linewidth]{img/YCSB-Domeinmodel.png}
\caption{YCSB: Domeinmodel YCSB in IMP}
\label{fig:imp-ycsb-domeinmodel}
\end{figure}

Dit model bevat maar 1 nieuw element en dit is YCSB. Deze dient verbonden te zijn met één van de 3 systemen die hierboven zijn beschreven. Elk van deze systemen zal getest worden voor alle testen die geactiveerd zijn. MongoDB heeft 2 connecties omdat de cluster voor de beschikbaarheidstesten wordt gebruikt en een replicaset voor de consistentie testen. Pgpool-II heeft geen ondersteuning voor de consistentie testen.

\subsection{Voorbeeld configuratie}

De configuratie voor de testomgeving gaat als volgt in IMP:

\lstinputlisting[language=Python, breaklines=true, frame=single]{code/imp-ycsb.conf}

De testen starten door het uitvoeren van {{directory}}/scripts/ycsb-script. De resultaten komen in de folder {{directory}}/results.
